Enhancing HiC data resolution with convolutional neural networks

### 2018-04-22 HiCPlus/tex/review.tex
almost finished report
done with the report

\maketitle
\section{Abstract}
\section{Introduction}
The cell of a eukaryotic species forms a multi-granularity genome structure
in order to compactly store a very long genomic 
DNA sequence in its small nucleus. The following describes the genome
structure in the order of decreasing granularity
\cite{wang2013properties}:
\begin{enumerate*}
\item[(1)] A \textbf{nucelotide} is the building block of
DNA. There are 4 types of nucleotides: 
C, G, A and T. 
\item[(2)] Each pair of nucleotides in the DNA are called a \textbf{base}.
A kilo-base is a group of 1000 bases.
\item[(3)] thousands of bases join together to form \textbf{gene loci}.
\item[(4)] A number of loci then fold into a large
independent physical structure called \textbf{chromosome}.
\end{enumerate*}

Study of spatial conformation of chromosomes is
of high importance in the field of (computational)
biology. Although all cell is a living being
have the same sequence of genes, it is the 
3D positioning of these genese in space that
determines how the cell functions.
Roughly said,
if two genes are close to each other in
space, they can interact with each other
in order to create a certain protein that
regulates a certain task.
Thus, being
able study this 3D configuration can help
unravel mysteries of cell functioning.
However, this spatial organization of chromosomes
can not be observed through traditional 
microscopy. As an alternative,
high-throughput chromosome conformation capture
(Hi-C) has emerged as
a powerfull method for studying the
3D organization of chromosomes in space.
In this method, a chromosome is divided into
very small equally sized
sections called \textit{loci}
which is composed of 1K to 1M bases.
this method then
measures all pair-wise interaction frequencies 
across all chromosomes. 
In the past years, Hi-C method has lead to some
exiting discoveries about the topology of 
chromosomes such as presence of \textit{chromatin
loops}.
Hi-C data are usually provided as a $N \times N$
heatmap or \textit{contact matrix} where $N$ is 
the number of loci in the genome. Each cell in 
the heatmap indicates the number of \textit{interactions}
found between a pair of loci corresponding to the
rows and columns. `Resolution' of a Hi-C data
is the size of the loci the genome is
divided into.
As mentioned above
resolution can range from 1 kb to 1 Mb.
\textit{sequencing depth} is the most
important factor that determines the resolution
of data. A higher sequencing depth results in
capturing interactions between samller loci,
thus improving the resolution of the data.
the sequencing process is costly and 
linewar increase of resolution requires
quadratic increase of sequencing reads.
thus, most of the Hi-C data availabe have
low resolutions.

@@

@@

\subsubsection{Loss Function}
We used mean squre of

@@

the model picks the middle 28 rows and colums of the
corresponing high-resolution patch and flattens
it. It then calculates the mean square of differneces

@@

We trained the model using low-resolution 
 GM06990 and high-resolution
 GM12878 data. We then used the other threed
 data that we have (corresponding to three
 cancerous cells) as input to the model to 
 improve contrast.
\section{Strengths and Weaknesses and Future Work}
This method can be considered as a noise reduction method
for Hi-C contact matrices. Just as all currently available
normalization and noise reduction approaches this method
also relies on assumptions. The major assumption of this
approach is that CNNs can predict values in high-resolution
matrix from the surrounding neighborhood in low-resolution
matrix. The difference of this assumption from other 
approaches is that it is easier to put the asumption to
statistical tests. Authors of \cite{zhang2018enhancing}
came to the conclusion that the assumption holds true for
a down-sampled version of high-resolution data. Whether
this is the case for an independent low-resolution data
remains to be investigated in future works.

>>>>

\maketitle
\section{Abstract}
\section{Introduction}
The cell of a eukaryotic species forms a multi-granularity genome structure
in order to compactly store a very long genomic 
DNA sequence in its small nucleus. 
A \textbf{nucelotide} is the building block of
DNA. There are 4 types of nucleotides: 
C, G, A and T. 
Each pair of nucleotides in the DNA are called a \textbf{base}.
A kilo-base is a group of 1000 bases.
thousands of bases join together to form \textbf{gene loci}.
A number of loci then fold into a large
independent physical structure called \textbf{chromosome}
        (\cite{wang2013properties}).

Study of spatial conformation of chromosomes is
of high importance in the field of (computational)
biology. Although all cell is a living being
have the same sequence of genes, it is the 
3D positioning of these genes in space that
determines how the cell functions.
Roughly said,
if two genes are close to each other in
space, they can interact with each other
in order to create a certain protein that
regulates a certain task.
Thus, being
able study this 3D configuration can help
unravel mysteries of cell functioning.
However, this spatial organization of chromosomes
can not be observed through traditional 
microscopy. As an alternative,
high-throughput chromosome conformation capture
(Hi-C) has emerged as
a powerful method for studying the
3D organization of chromosomes in space.
The HiC method, which was developed by 
\cite{lieberman2009comprehensive}, 
captures interactions between
chromosomal fragments.
In this method, a chromosome is divided into
very small equally sized
sections called \textit{loci}
which is composed of 1K to 1M bases.
this method then
measures all pair-wise interaction frequencies 
across all chromosomes. 
In the past years, Hi-C method has lead to some
exiting discoveries about the topology of 
chromosomes such as presence of \textit{chromatin
loops}.
Hi-C data are usually provided as a $N \times N$
heatmap or \textit{contact matrix} where $N$ is 
the number of loci in the genome. Each cell in 
the heatmap indicates the number of \textit{interactions}
found between a pair of loci corresponding to the
rows and columns. `Resolution' of a Hi-C data
is the size of the loci the genome is
divided into.
As mentioned above
resolution can range from 1 kb to 1 Mb.
\textit{sequencing depth} is the most
important factor that determines the resolution
of data. A higher sequencing depth results in
capturing interactions between smaller loci,
thus improving the resolution of the data.
the sequencing process is costly and 
linear increase of resolution requires
quadratic increase of sequencing reads.
thus, most of the Hi-C data available have
low resolutions.

@@

As to malignant cells,
we re-used Leukemic Hi-C libraries 
created in \cite{wang2013properties}
These libraries were sequenced 
for cases of primary human B-acute
lymphoblastic leukemia (B-ALL or ALL), 
the MHH-CALL-4 B-ALL cell
line (CALL4), 
and the follicular lymphoma cell-line (RL).
Just as \cite{wang2013properties}, 
we used normal B-cell line (GM068990)
from \cite{lieberman2009comprehensive} 
for our comparisons.
    \begin{tabular}{llcl}
        Data                           & Short & Resolution & Used …                     \\ \hline\hline\\
        Normal B-cell(GM12878)         & RAO   & high       & As response in training    \\[5pt]\\
        Normal B-cell(GM06990)         & MIT   & low        & As feature in training     \\[5pt]\\
        B-acute lymphoblastic leukemia & ALL   & low        & To predict high resolution \\[5pt]\\
        Follicular lymphoma cell-line  & RL    & low        & To predict high resolution \\[5pt]\\
        MHH-CALL-4                     & CALL4 & low        & To predict high resolution \\[10pt]
    \caption{Data description}
    \label{tab:data}

@@

\subsection{Loss Function}
We used mean square of

@@

the model picks the middle 28 rows and columns of the
corresponding high-resolution patch and flattens
it. It then calculates the mean square of differeneces

@@

    \includegraphics[width=\textwidth]{mit.png}
    \caption{}
        Malignant Cell 1: B-ALL
        \includegraphics[width=\textwidth]{all.png}
        \vspace{10pt}
        Malignant Cell 2: RL
        \includegraphics[width=\textwidth]{rl.png}
        \vspace{10pt}
        Malignant Cell 3: CALL4
        \includegraphics[width=\textwidth]{call4.png}
    \caption{}

\textbf{Platforms:}
The original model was developed in \texttt{theano} 
framework.
The package \texttt{lasagne} was used 
for model development. Although we made some experiments
with tensorflow, conducted our final traning via the
orginal model.
Our results can be found in a git repository (refer to
resources section).

We trained the model using low-resolution 
 GM06990 and high-resolution
 GM12878 data, both of which are sequence from
 the same cell. We then used the other three
 data that we have (corresponding to three
 cancerous cells) as input to the model to 
 improve contrast.
The data that we uses is described in table
\ref{tab:data}
\section{Strengths and Weaknesses and Future Work}
This method can be considered as a noise reduction method
for Hi-C contact matrices. 
For example Balance Network Deconvolution,
proposed by \cite{feizi2013network}
assumes that the observed graph $G_{obs}$ is a summation of its
direct graph $G_{dir}$ and some indirect terms as follows:
\begin{equation}
        \label{indirect_effects}
            G_{obs} = G_{dir} + G_{dir}^2 + G_{dir} + ...
\end{equation}
Just as all currently available
normalization and noise reduction approaches this method
also relies on assumptions. The major assumption of this
approach is that CNNs can predict values in high-resolution
matrix from the surrounding neighborhood in low-resolution
matrix. The difference of this assumption from other 
approaches is that it is easier to put the assumption to
statistical tests. Authors of \cite{zhang2018enhancing}
came to the conclusion that the assumption holds true for
a down-sampled version of high-resolution data. Whether
this is the case for an independent low-resolution data
remains to be investigated in future works.
\section{Resources}
\textbf{Source code for the project:}\\
\url{https://github.com/rasoolianbehnam/watson/tree/master/HiCPlus}\\
\textbf{Link to the low-resolution Hi-C datasets:}\\
\url{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/contact_file/}\\
\textbf{Link to high-resolution Hi-C dataset:}\\
\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63525}
